---
title: "Assignment 04"
author: "Dan Hadley"
date: '2019-10-08'
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "hw04_files/")

## Keep track of list of tables
tab.cap = c('Table 1: Minimum and Maximum Country Level GDP per Capita by Continent and the corresponding countries and years',
            'Table 2: Minimum and Maximum Continent Level GDP per Capita and the corresponding year',
            'Table 3: Summary statistics for the Distribution of Country GDP per Capita by Continent',
            'Table 4: Life expectancies (in years) using population-weighted means across all countries globally and within continents')

fig.cap = c('Figure 1: Scatterplot of life expectancies between Canada and the United States',
            'Figure 2: Jittered plot of country-level GDP per capita grouped by continent',
            'Figure 3: Global life expectancy and continent-level life expectancies over time')
```

## Introduction

For this analysis, we load the `gapminder`, `tidyverse`, `tidyr` and `DT` packages.  The code to load these packages is not displayed in the report.  Data wrangling is performed on the gapminder dataset.  In this report, I have chosen the following:

  * [__Univariate Option 1__](#exercise-1---univariate-option-1) - Using the gapminder dataset, we make a new tibble with one row per year and two columns containing the life expectancies for two countries, use this new tibble to make a scatterplot of the life expectancies, and then re-lengthen the data.
  * [__Multivariate Option 1__](#exercise-2---multivariate-option-1) - Using the gapminder dataset, we make a tibble with one row per year and columns for both life expectancies and GDP per capita for two countries and then re-lengthen the data.
  * [__Exercise 3__](#exercise-3---table-joins) - 

```{r load packages, warning=FALSE, message=FALSE}
suppressPackageStartupMessages(library(gapminder)) # for gapminder dataset
suppressPackageStartupMessages(library(tidyverse)) #loads ggplot2, tibble, tidyr, dplyr packages
suppressPackageStartupMessages(library(DT)) #for datatable() command to display datatables
```

## Exercise 1 - Univariate Option 1

### Make a Tibble

Observations in the `gapminder::gapminder` dataset can be defined as country/year.  Thus, the dataset can be considered tidy since each row corresponds to one country/year observation.  In this section, we make a tibble with one row per year and columns for life expectancy for two or more countries.  The countries used are Canada and United States.

```{r}
can.us <- gapminder %>% 
  #filter to Canada and United States
  filter(country == "Canada" | country == "United States") %>% 
  #select year, country, and life expectancy columns
  select(year, country, lifeExp) %>% 
  # Transform tibble with year, country, and lifeExp columns to year, Canada and United States with life expectancies
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp)

datatable(can.us)
```

### Scatterplot

In this section, we use the new tibble `can.us` to produce a scatterplot of the life expectancy of Canada versus the life expectancy of the United States.  Note that since the column "United States" has a space in the middle, the code uses backticks, \`, so that the aesthetics are properly mapped.

```{r lifeExp, fig.cap=fig.cap[1]}
can.us %>% 
  ggplot(aes(Canada, `United States`)) +
  geom_point() +
  ggtitle(label='Life Expectancies (years)')
```

### Re-lengthen 

In this step, the `can.us` tibble is re-lengthened.

```{r}
can.us %>% 
  pivot_longer(cols = c("Canada",
                        "United States"),
               names_to = "country",
               values_to = "lifeExp") %>% 
  arrange(country) %>% 
  datatable()
```

## Exercise 2 - Multivariate Option 1

### Make a Tibble

In this section, we make a tibble with one row per year and columns for both life expectancy and GDP per capita for two unique countries.  The countries chosen are China and India.  Both life expectancy and GDP per capita are rounded to two decimal places.

```{r}
ch.in <- gapminder %>% 
  filter(country == "China" |
           country == "India") %>% 
  select(country, year, lifeExp, gdpPercap) %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap)) %>% 
  mutate_at(vars(-year), ~round(., 2))

datatable(ch.in)
```

### Re-lengthen

In this section, we re-lengthen the data so that each row's observation is defined as the year/country intersection.  Since we have 2 countries and `r length(unique(ch.in$year))` years, the re-lengthened data has a total of `r 2*length(unique(ch.in$year))` rows instead of the `r nrow(ch.in)` rows as in the tibble from the previous section.

```{r}
ch.in %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_") %>% 
  datatable()
```
As you can see from the above tibble, we've recreated the long-form data set that we originally started with!

## Exercise 3 - Table Joins

Begin by reading in guest and email.

```{r}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")

```

### 3.1 
For each guest in the guestlist tibble, add a column for email address, which can be found in email tibble.

First, separate the emails by row.
```{r}
sep_emails <- email %>% 
  separate_rows(guest, sep = ", ")

datatable(sep_emails)
```

```{r}
joined_list <- guest %>% 
  rename(guest = "name") %>% 
  left_join(sep_emails, by="guest")

datatable(joined_list)
```

### 3.2
We can move on to the next part: figuring out who we have emails for, but they're not yet on the guestlist. To do this, we will need to look at the email list, then compare and see who's not on the main guestlist.

```{r}
not_in_guestlist <- sep_emails %>%  
  rename(name = "guest") %>% 
  anti_join(guest, by="name")

datatable(not_in_guestlist)
```

As you can see, using the anti-join function allows us to produce all the people who are in the email data frame (i.e: they have an email), but are not included in g (i.e: the guestlist). 

### 3.3 
In this section, we will make a guestlist that includes everyone we have emails for (in addition to the onles in the original guestlist). To do this, we will simply join the table we made in 3.2 with the original guestlist table but appending them to the end. 

```{r}
guest.all <- joined_list %>% 
  rename(name = "guest") %>% 
  bind_rows(not_in_guestlist)
datatable(guest.all)
```

The above guestlist contains everyone from the original guestlist, plus all the ones that we have emails for but were not in the original guestlist. Obviously, the new additions do not have information regarding their attendance. 

