---
title: "Assignment 04"
author: "Dan Hadley"
date: '2019-10-08'
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "hw04_files/")

## Keep track of list of tables
tab.cap = c('Table 1: Life expectancies (years) for Canada and the United States',
            'Table 2: Re-lengthened dataset for life expectancies for Canada and the United States',
            'Table 3: Life expectancies (years) and GDP per capitas ($US-inflation adjusted) for China and India',
            'Table 4: Re-lengthed dataset from Table 3')

fig.cap = c('Figure 1: Scatterplot of life expectancies between Canada and the United States')
```

## Introduction

For this analysis, we load the `gapminder`, `tidyverse`, `tidyr` and `DT` packages.  The code to load these packages is not displayed in the report.  Data wrangling is performed on the gapminder dataset.  In this report, I have chosen the following:

  * [__Univariate Option 1__](#exercise-1---univariate-option-1) - Using the gapminder dataset, we make a new tibble with one row per year and two columns containing the life expectancies for two countries, use this new tibble to make a scatterplot of the life expectancies, and then re-lengthen the data.
  * [__Multivariate Option 1__](#exercise-2---multivariate-option-1) - Using the gapminder dataset, we make a tibble with one row per year and columns for both life expectancies and GDP per capita for two countries and then re-lengthen the data.
  * [__Exercise 3__](#exercise-3---table-joins) - 

```{r load packages, warning=FALSE, message=FALSE}
suppressPackageStartupMessages(library(gapminder)) # for gapminder dataset
suppressPackageStartupMessages(library(tidyverse)) #loads ggplot2, tibble, tidyr, dplyr packages
suppressPackageStartupMessages(library(DT)) #for datatable() command to display datatables
```

## Exercise 1 - Univariate Option 1

### Make a Tibble

Observations in the `gapminder::gapminder` dataset can be defined as country/year.  Thus, the dataset can be considered tidy since each row corresponds to one country/year observation.  In this section, we make a tibble with one row per year and columns for life expectancy for two or more countries.  The countries used are Canada and United States.

```{r Q1.1}
can.us <- gapminder %>% 
  #filter to Canada and United States
  filter(country == "Canada" | country == "United States") %>% 
  #select year, country, and life expectancy columns
  select(year, country, lifeExp) %>% 
  # Transform tibble with year, country, and lifeExp columns to year, Canada and United States with life expectancies
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp)

datatable(can.us, caption=tab.cap[1])
```

### Scatterplot

In this section, we use the new tibble `can.us` to produce a scatterplot of the life expectancy of Canada versus the life expectancy of the United States.  Note that since the column "United States" has a space in the middle, the code uses backticks, \`, so that the aesthetics are properly mapped.

```{r Q1.2, fig.cap=fig.cap[1]}
can.us %>% 
  ggplot(aes(Canada, `United States`)) +
  geom_point() +
  ggtitle(label='Life Expectancies (years)')
```

### Re-lengthen 

In this step, the `can.us` tibble is re-lengthened.

```{r Q1.3}
can.us %>% 
  pivot_longer(cols = c("Canada",
                        "United States"),
               names_to = "country",
               values_to = "lifeExp") %>% 
  arrange(country) %>% 
  datatable(caption=tab.cap[2])
```

## Exercise 2 - Multivariate Option 1

### Make a Tibble

In this section, we make a tibble with one row per year and columns for both life expectancy and GDP per capita for two unique countries.  The countries chosen are China and India.  Both life expectancy and GDP per capita are rounded to two decimal places.

```{r Q2.1}
ch.in <- gapminder %>% 
  filter(country == "China" |
           country == "India") %>% 
  select(country, year, lifeExp, gdpPercap) %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap)) %>% 
  mutate_at(vars(-year), ~round(., 2))

datatable(ch.in, caption=tab.cap[3])
```

### Re-lengthen

In this section, we re-lengthen the data so that each row's observation is defined as the year/country intersection.  Since we have 2 countries and `r length(unique(ch.in$year))` years, the re-lengthened data has a total of `r 2*length(unique(ch.in$year))` rows instead of the `r nrow(ch.in)` rows as in the tibble from the previous section.

```{r Q2.2}
ch.in %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_") %>% 
  datatable(caption=tab.cap[4])
```
As you can see from the above tibble, we've recreated the long-form data set that we originally started with!

## Exercise 3 - Table Joins

We begin by importing made-up wedding guestlist and e-mail addresses using the `read_csv()` function from the `readr` package which is automatically loaded with the `tidyverse` package.

```{r readin}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")

```

### Add E-mail Column

For each guest in the guestlist, which is the tibble `guest`, we add a column for the e-mail address.

First, separate the emails by row.
```{r}
sep_emails <- email %>% 
  separate_rows(guest, sep = ", ")

datatable(sep_emails)
```

```{r}
joined_list <- guest %>% 
  rename(guest = "name") %>% 
  left_join(sep_emails, by="guest")

datatable(joined_list)
```

### E-mails not on Guestlist

In this section, we find e-mails in our e-mail list (tibble `email`) for persons not on the guestlist.

```{r}
not_in_guestlist <- sep_emails %>%  
  rename(name = "guest") %>% 
  anti_join(guest, by="name")

datatable(not_in_guestlist)
```

As you can see, using the anti-join function allows us to produce all the people who are in the email data frame (i.e: they have an email), but are not included in g (i.e: the guestlist). 

### Expand Guestlist
In this section, we make a guestlist that includes everyone on the original guestlist and everyone for whom we have emails. This is accomplished by joining the datasets `guest` and `not_in_guestlist` created in the previous section. 

```{r}
guest.all <- joined_list %>% 
  rename(name = "guest") %>% 
  bind_rows(not_in_guestlist)
datatable(guest.all)
```


