---
title: "Assignment 04"
author: "Dan Hadley"
date: '2019-10-08'
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "hw04_files/")

## Keep track of list of tables
tab.cap = c('Table 1: Life expectancies (years) for Canada and the United States',
            'Table 2: Re-lengthened dataset for life expectancies for Canada and the United States',
            'Table 3: Life expectancies (years) and GDP per capitas ($US-inflation adjusted) for China and India',
            'Table 4: Re-lengthed dataset from Table 3',
            'Table 5: Guest list with e-mails where available',
            'Table 6: Guests from the guest list with no e-mail available',
            'Table 7: Names of people with e-mails available that are not on the guest list',
            'Table 8: Combined tibble of all people either on the guest list or the e-mail list')

fig.cap = c('Figure 1: Scatterplot of life expectancies between Canada and the United States')
```

## Introduction

For this analysis, we load the `gapminder`, `tidyverse`, `knitrr` and `DT` packages.  The code to load these packages is not displayed in the report.  Data wrangling is performed on the gapminder dataset.  In this report, I have chosen the following:

  * [__Univariate Option 1__](#exercise-1---univariate-option-1):
    - Make a new tibble with one row per year, and columns for the life expectancy for two countries
    - Take advantage of this new data shape to scatterplot life expectancy for one country against that of another
    - Re-lengthen the data
  * [__Multivariate Option 1__](#exercise-2---multivariate-option-1):
    - Make a tibble with one row per year and columns for both life expectancies and GDP per capita for two countries
    - Re-lengthen the data
  * [__Exercise 3__](#exercise-3---table-joins):
    - For each guest in the guestlist (`guest` tibble), add a column for email address which can be found in the `email` tibble
    - Who do we have emails for, yet are not on the guestlist
    - Make a guest list that includes everyone we have emails for (in addition to those on the original guestlist)

```{r load packages, warning=FALSE, message=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(gapminder)) # for gapminder dataset
suppressPackageStartupMessages(library(tidyverse)) #loads ggplot2, tibble, tidyr, dplyr packages
suppressPackageStartupMessages(library(DT)) #for datatable() command to display datatables
suppressPackageStartupMessages(library(knitr)) #for kable() command
```

## Exercise 1 - Univariate Option 1

### Make a Tibble

Observations in the `gapminder::gapminder` dataset can be defined as country/year.  Thus, the dataset can be considered tidy since each row corresponds to one country/year observation.  In this section, we make a tibble with one row per year and columns for life expectancy for two or more countries.  The countries used are Canada and United States.

```{r Q1.1}
can.us <- gapminder %>% 
  #filter to Canada and United States
  filter(country == "Canada" | country == "United States") %>% 
  #select year, country, and life expectancy columns
  select(year, country, lifeExp) %>% 
  # Transform tibble with year, country, and lifeExp columns to year, Canada and United States with life expectancies
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp)

datatable(can.us, caption=tab.cap[1])
```

### Scatterplot

In this section, we use the new tibble `can.us` to produce a scatterplot of the life expectancy of Canada versus the life expectancy of the United States.  Note that since the column "United States" has a space in the middle, the code uses backticks, \`, so that the aesthetics are properly mapped.

```{r Q1.2, fig.cap=fig.cap[1]}
can.us %>% 
  ggplot(aes(Canada, `United States`)) +
  geom_point() +
  ggtitle(label='Life Expectancies (years)')
```

### Re-lengthen 

In this step, the `can.us` tibble is re-lengthened.

```{r Q1.3}
can.us %>% 
  pivot_longer(cols = c("Canada",
                        "United States"),
               names_to = "country",
               values_to = "lifeExp") %>% 
  arrange(country) %>% 
  datatable(caption=tab.cap[2])
```

## Exercise 2 - Multivariate Option 1

### Make a Tibble

In this section, we make a tibble with one row per year and columns for both life expectancy and GDP per capita for two unique countries.  The countries chosen are China and India.  Both life expectancy and GDP per capita are rounded to two decimal places.

```{r Q2.1}
ch.in <- gapminder %>% 
  filter(country == "China" |
           country == "India") %>% 
  select(country, year, lifeExp, gdpPercap) %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap)) %>% 
  mutate_at(vars(-year), ~round(., 2))

datatable(ch.in, caption=tab.cap[3])
```

### Re-lengthen

In this section, we re-lengthen the data so that each row's observation is defined as the year/country intersection.  Since we have 2 countries and `r length(unique(ch.in$year))` years, the re-lengthened data has a total of `r 2*length(unique(ch.in$year))` rows instead of the `r nrow(ch.in)` rows as in the tibble from the previous section.

```{r Q2.2}
ch.in %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_") %>% 
  datatable(caption=tab.cap[4])
```
As you can see from the above tibble, we've recreated the long-form data set that we originally started with!

## Exercise 3 - Table Joins

We begin by importing made-up wedding guestlist and e-mail addresses using the `read_csv()` function from the `readr` package which is automatically loaded with the `tidyverse` package.

```{r read-in, message=FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

### 3.1 - Add E-mail Column

For each guest in the guestlist (tibble `guest`), we add a column for the e-mail address.  We do this by first separating rows by `, `, since the names of all guests from the same party are on the same line separated by a comma in the `email` tibble.  The `separate_rows()` function is designed for this specific purpose. Once we modify the `email` tibble in this way, we can compare it to `guest` tibble using the `right_join()` function.


```{r Q3.1}
#First separate rows in email tibble to get one guest per row
guest.email <- email %>% 
  separate_rows(guest, sep = ", ") %>% 
  right_join(guest, by=c("guest" = "name"))

datatable(guest.email, caption=tab.cap[5])
```

We see that there are `r summarize(guest.email, sum(is.na(email)))` guests that do not have an e-mail address in the `email` tibble.  These guests are `r filter(guest.email, is.na(email)) %>% select(guest) %>% kable(caption=tab.cap[6])`

### 3.2 - E-mails not on Guestlist

In this section, we find e-mails in our e-mail list (tibble `email`) for persons not on the guestlist.  This is similar to Exercise 3.1, but instead of using the `left_join()` or `right_join()` functions, we use `anti_join()`, which returns all rows from a tibble that do not match values in another tibble.

```{r}
email.only <- email %>%
  separate_rows(guest, sep=", ") %>% 
  anti_join(guest, by=c("guest" = "name"))

datatable(email.only, caption=tab.cap[7])
```


### 3.3 - Expand Guestlist
In this section, we combine the guest list with the e-mail list so that each name appears exactly once in a new tibble.

```{r}
guest.email %>%
  bind_rows(email.only) %>% 
datatable(caption=tab.cap[8])
```


