---
title: "Assignment 04"
author: "Dan Hadley"
date: '2019-10-08'
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "hw04_files/")

## Keep track of list of tables
tab.cap = c('Table 1: Life expectancies (years) for Canada and the United States',
            'Table 2: Re-lengthened dataset of life expectancies for Canada and the United States',
            'Table 3: Life expectancies (years) and GDP per capitas ($US-inflation adjusted) for China and India',
            'Table 4: Re-lengthed dataset from Table 3',
            'Table 5: Guest list with e-mails where available',
            'Table 6: Guests from the guest list with no e-mail available',
            'Table 7: Names of people with e-mails available that are not on the guest list',
            'Table 8: Combined tibble of all people either on the guest list or the e-mail list')

fig.cap = c('Figure 1: Scatterplot of life expectancies between Canada and the United States with larger points for more recent observations')
```

## Introduction

For this analysis, we load the `gapminder`, `tidyverse`, `knitrr` and `DT` packages.  The code to load these packages is not displayed in the report.  Data wrangling is performed on the gapminder dataset.  In this report, I have chosen the following:

  * [__Univariate Option 1__](#exercise-1---univariate-option-1):
    - Make a new tibble with one row per year, and columns for the life expectancy for two countries
    - Take advantage of this new data shape to scatterplot life expectancy for one country against that of another
    - Re-lengthen the data
  * [__Multivariate Option 1__](#exercise-2---multivariate-option-1):
    - Make a tibble with one row per year and columns for both life expectancies and GDP per capita for two countries
    - Re-lengthen the data
  * [__Exercise 3__](#exercise-3---table-joins):
    - For each guest in the `guest` tibble, add a column for email address which can be found in the `email` tibble
    - Who do we have emails for, yet are not on the guestlist
    - Make a guest list that includes everyone we have emails for (in addition to those on the original guestlist)

```{r load packages, warning=FALSE, message=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(gapminder)) # for gapminder dataset
suppressPackageStartupMessages(library(tidyverse)) #loads ggplot2, tibble, tidyr, dplyr packages
suppressPackageStartupMessages(library(DT)) #for datatable() command to display datatables
suppressPackageStartupMessages(library(knitr)) #for kable() command
```

## Exercise 1 - Univariate Option 1

### Make a Tibble

Observations in the `gapminder::gapminder` dataset can be defined as country/year.  Thus, the dataset can be considered tidy since each row corresponds to one country/year observation.  In this section, we make a tibble with one row per year and columns for life expectancy for Canada and United States.  The data is presented in Table 1.

```{r Q1.1}
can.us <- gapminder %>% 
  filter(country == "Canada" | country == "United States") %>% #filter by country
  select(year, country, lifeExp) %>% #select columns of interest
  pivot_wider(id_cols = year, #observations
              names_from = country, #new column names
              values_from = lifeExp) #new column values

datatable(can.us, caption=tab.cap[1])
```

### Scatterplot

In this section, we use the new tibble `can.us` to produce a scatterplot of the life expectancy of Canada versus the life expectancy of the United States.  Note that since the column name "United States" has a space, the code uses backticks, \`United States\`, so that the aesthetics are properly mapped.

```{r Q1.2, fig.cap=fig.cap[1]}
can.us %>% 
  ggplot(aes(Canada, `United States`)) +
  geom_point(shape=1, aes(size=year)) +
  ggtitle(label='Scatterplot of Life Expectancies (years) for Two Countries')
```

### Re-lengthen 

The `can.us` tibble is re-lengthened so that observations are year/country combinations and presented as Table 2.

```{r Q1.3}
can.us %>% 
  pivot_longer(cols = c("Canada", "United States"),
               names_to = "country",
               values_to = "lifeExp") %>% 
  arrange(country) %>% 
  datatable(caption=tab.cap[2])
```

## Exercise 2 - Multivariate Option 1

### Make a Tibble

In this section, we make a tibble with one row per year and columns for both life expectancy and GDP per capita for two countries, China and India and presented in Table 3.  Separating the data and country by underscore in the column names leads to easier re-lengthening in the next section.

```{r Q2.1}
ch.in <- gapminder %>% 
  filter(country == "China" | country == "India") %>% #filter by country
  select(country, year, lifeExp, gdpPercap) %>% #columns of interest
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap)) %>% 
  mutate_at(vars(-year), ~round(., 2)) #round numeric data

datatable(ch.in, caption=tab.cap[3])
```

### Re-lengthen

In this section, we re-lengthen the data so that an observation is defined as a year/country intersection.  The data is presented in Table 4.  Since we have 2 countries and `r length(unique(ch.in$year))` years, Table 4 has `r 2*length(unique(ch.in$year))` rows instead of the `r nrow(ch.in)` rows of Table 3.

```{r Q2.2}
ch.in %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_") %>% 
  arrange(country) %>% 
  datatable(caption=tab.cap[4])
```

## Exercise 3 - Table Joins

We begin by importing made-up wedding guestlist and e-mail addresses using the `read_csv()` function from the `readr` package which is automatically loaded with the `tidyverse` package.

```{r read-in, message=FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

### 3.1 - Add E-mail Column

For each guest in the guestlist (tibble `guest`), we add a column for the e-mail address.  The `separate_rows()` function is easy to use here due to the formatting of the names in the "guest" column of the `email` tibble. Then, we can compare separated rows to the `guest` tibble using the `right_join()` function.  The results are presented in Table 5.

```{r Q3.1}
#First separate rows in email tibble to get one guest per row
guest.email <- email %>% 
  separate_rows(guest, sep = ", ") %>% #create row per guest name
  right_join(guest, by=c("guest" = "name")) #adds emails to guest tibble

datatable(guest.email, caption=tab.cap[5])
```

We see that there are `r summarize(guest.email, sum(is.na(email)))` guests that do not have an e-mail address in the `email` tibble.  These guests are `r filter(guest.email, is.na(email)) %>% select(guest) %>% kable(caption=tab.cap[6])`

### 3.2 - E-mails not on Guestlist

In this section, we find e-mails in the `email` tibble for persons not on the guestlist.  Using `anti_join()`, this list is created and presented in Table 7.

```{r}
email.only <- email %>%
  separate_rows(guest, sep=", ") %>% #create one row per person in email tibble
  anti_join(guest, by=c("guest" = "name")) #compare email$guest to guest$name

datatable(email.only, caption=tab.cap[7])
```


### 3.3 - Expand Guestlist
In this section, we combine the guest list with the e-mail list so that each name appears exactly once in a new tibble.

```{r}
guest.email %>%
  bind_rows(email.only) %>% 
datatable(caption=tab.cap[8])
```


