---
title: 'Homework 02'
author: 'Dan Hadley'
output: 
  html_document:
    keep_md: true
    toc: true
editor_options: 
  chunk_output_type: inline
---
<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE,
                      fig.path = "gapminder_files/")
```

**Optional, but recommended startup**:

1. Change the file output to both html and md _documents_ (not notebook).

2. `knit` the document. 

3. Stage and commit the rmd, and knitted documents.

# Intro to `dplyr` syntax

Load the `gapminder` and `tidyverse` packages. Hint: `suppressPackageStartupMessages()`!
    - This loads `dplyr`, too.

<!---The following chunk loads the appropriate libraries--->

```{r load packages, echo=FALSE, warning = FALSE, message = FALSE}
# load your packages here:
suppressPackageStartupMessages(library(gapminder)) #to use the tibble dataset gapminder
suppressPackageStartupMessages(library(tidyverse)) #to load in dplyr package
suppressPackageStartupMessages(library(DT)) #to nicely display tibbles with datatable()
suppressPackageStartupMessages(library(knitr)) #display tables in HTML for R output with kable()
suppressPackageStartupMessages(library(gridExtra)) #change the layout of multiple ggplots()
```
    
# Question 1

## 1.1 Subset to Three Countries in the 1970's

<!---The following chunk finds the first year of observations for the 70's and the top 3 countries by population for that year--->

```{r subset, echo=FALSE}
## Create data for observations in the 1970's
gap.70s <- gapminder %>% 
  filter(year > 1969,
         year < 1980)

## Create list of years in the 1970's for which there are observations
yrs <- gap.70s %>% 
  select(year) %>% 
  unique()

## Find countries with largest populations as of first year of observations during 1970's
n.ctries <- 3 #number of countries to select
ctries <- gap.70s %>% # Pipe command from the dplyr package can be added with keyboard shortcut command+shift+m
  filter(year == min(yrs)) %>% # Filter to observations from 1972
  arrange(desc(pop)) %>% # Arrange by population
  slice(1:n.ctries) %>% # Choose top n.ctries rows
  select(country) %>% # Select country column
  .$country %>% # Select column
  as.character() # Convert to character vector from factor

```

Using the `filter()` function from the `dplyr` package, we filter the gapminder data to all observations for three countries in the 1970's.  To pick which three countries, we choose countries with the largest populations as of `r min(yrs)`: `r paste(ctries, collapse=", ")`.

```{r displaytibble}
## Create data with 3 countries' observations in the 1970's based on largest populations from first 1970's observation
(gap.3 <- gapminder %>% 
   filter(year > 1969,
          year < 1980,
          country %in% ctries))
```

## 1.2 Use Pipe Operator to Select Country and GDP per Capita:

The `select()` function from the `dplyr` package makes it very easy to choose which columns to display from a dataset.

```{r pipe}
## Select country and gdpPercap columns from filtered dataset
gap.3 %>% 
  select(country, gdpPercap)
```

## 1.3 Filter `gapminder` to observations with drop in Life Expectancy:

First, we can use the `mutate()` function to add a column to the `gapminder` dataset that calculates the 5-year change in life expectancy, in years, for each country.  This column is named `changeinlifeExp`.  For each country's first year observation, the change is 0.  Using this new column, we can filter out observations whose 5-year life expectancy change is negative.  When displaying the negative life expectancy changes, the change values are rounded to two decimal places.  Thus, a change that displays `0` is actually a negative change smaller than `-0.005`.

```{r LifeExp}
gapminder %>% 
  group_by(country) %>% 
  mutate(changeinlifeExp = c(0, diff(lifeExp, lag=1))) %>% 
  filter(changeinlifeExp < 0) %>% 
  mutate(changeinlifeExp = round(changeinlifeExp, 2)) %>% #round change to 2 digits
  datatable()
```

## 1.4 Show maximum Country GDP per Capita for each Country

```{r maxGDP}
gapminder %>% 
  group_by(country) %>% 
  summarise(maxgdp = round(max(gdpPercap), 2)) %>%
  datatable()
```

## 1.5 Scatterplot of Canada's Life Expectancy vs. GDP per Capita

```{r scatterplot, fig.width=8, fig.height=4}
gapminder %>% 
  filter(country == 'Canada') %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point() + 
  ggtitle('Canada') + 
  scale_x_log10('GDP per Capita', labels=scales::dollar_format()) + 
  ylab('Life Expectancy')
```

# Question 2

We analyze the `continent` categorical variable, listing the continent for each observation, and the `lifeExp` quantitative varible, which gives each observation's life expectancy in years.

A bar plot for continent shows the possible values of the `continent` variable as well as giving a sense of the spread of the observations by showing which continents are most/least observed, etc.

```{r continent}
ggplot(gapminder, aes(continent)) + 
  geom_bar() + 
  ggtitle('Count of Observations by Continent') + 
  theme(plot.title = element_text(color='black', face="bold")) + 
  ylab('Number of Observations') +
  xlab('Continent')
```

For life expectancy, 
```{r lifeExp, fig.width=10, fig.height=3}
plot1 <- ggplot(gapminder, aes(lifeExp)) + 
  geom_histogram(bins=50) + 
  ggtitle('Histogram of Life Expectancies') + 
  xlab('Life Expectancy') + 
  ylab('Number of Observations') + 
  theme(plot.title = element_text(color='black', face="bold"))

plot2 <- ggplot(gapminder, aes(y=lifeExp)) + 
  geom_boxplot() + 
  ggtitle('Boxplot of Life Expectancies') + 
  ylab('Life Expectancy') + 
  xlab('') + 
  theme(plot.title = element_text(color='black', face="bold"), 
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot3 <- ggplot(gapminder, aes(lifeExp)) + 
  geom_density() + 
  ggtitle('Density of Life Expectancies') + 
  ylab('Density') + 
  xlab('Life Expectancy') + 
  theme(plot.title = element_text(color='black', face="bold"))

grid.arrange(plot1, plot2, plot3, nrow=1, ncol=3)

summ.le <- summary(gapminder$lifeExp)
names(summ.le) <- c('Minimum', 'First Quartile', 'Median', 'Mean', 'Third Quartile', 'Maximum')
summ.le <- data.frame(rbind(round(summ.le, 2)), row.names = 'Life Expectancy')
kable(summ.le)

```

# Question 3

We now make two plots with some value for anlaysis.  The first plot is a scatterplot of GDP per capita by continent versus population-weighted mean life expectancy.  To calculate GDP per capita, we multiply each country in a continent by its popuSince we took a look at both continent and life expectancy in Question 2, we may want to view life expectancies by continent.  In order to compare the distribution of life expectancies across continents, we look at a boxplot.

```{r Q3, fig.width=8, fig.height=4}

## Population by Continent/Year
gapminder %>% 
  group_by(continent, year) %>% 
  mutate(wgt.lifeExp=weighted.mean(lifeExp, w=pop)) %>% 
  mutate(tot.pop=as.numeric(sum(pop)), tot.gdp=as.numeric(sum(pop*gdpPercap))) %>%  #as.numeric() needed for precision
  mutate(tot.gdpPer=round(tot.gdp/tot.pop, 2)) %>% #GDP per capita by continent
  ggplot(aes(x=tot.gdpPer, y=wgt.lifeExp)) + 
  geom_point(aes(shape=continent, color=continent, size=year)) + 
  scale_x_continuous('Continent GDP per Capita', labels=scales::dollar_format()) + 
  ylab('Populated Weighted Mean Life Expectancy') + 
  ggtitle('Continent Level GDP vs Life Expectancy') + 
  theme(plot.title = element_text(color='black', face="bold"))
```

```{r plot2, fig.width=9, fig.height=4}
plot1 <- ggplot(gapminder, aes(x=continent, y=lifeExp)) + 
  geom_boxplot() + 
  xlab('Continent') + 
  ylab('Life Expectancy') + 
  ggtitle('Life Expectancy by Continent') + 
  theme(plot.title = element_text(color='black', face="bold"))

plot2 <- gapminder %>% 
  group_by(continent) %>% 
  mutate(Result=n_distinct(country)) %>% 
  select(continent, Result) %>% 
  unique() %>% 
  ggplot(aes(x=continent, y=Result)) +
  geom_col() +
  ggtitle('Number of Countries from Each Continent') + 
  ylab('Number of Unique Countries') +
  xlab('Continent') + 
  theme(plot.title = element_text(color='black', face="bold"))

grid.arrange(plot1, plot2, nrow=1, ncol=2)
```
