---
title: 'Homework 02'
author: 'Dan Hadley'
output: 
  github_document:
    toc: true
---
<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE,
                      fig.path = "gapminder_files/")
```

<!---The following chunk loads the appropriate libraries--->

```{r load packages, echo=FALSE, warning = FALSE, message = FALSE}
# load your packages here:
suppressPackageStartupMessages(library(gapminder)) #to use the tibble dataset gapminder
suppressPackageStartupMessages(library(tidyverse)) #to load in dplyr package
suppressPackageStartupMessages(library(DT)) #to nicely display tibbles with datatable()
suppressPackageStartupMessages(library(knitr)) #display tables in HTML for R output with kable()
suppressPackageStartupMessages(library(gridExtra)) #change the layout of multiple ggplots()
```
    
## Question 1

### 1.1 Subset to Three Countries in the 1970's

<!---The following chunk finds the first year of observations for the 70's and the top 3 countries by population for that year--->

```{r subset, echo=FALSE}
## Create data for observations in the 1970's
gap.70s <- gapminder %>% 
  filter(year > 1969,
         year < 1980)

## Create list of years in the 1970's for which there are observations
yrs <- gap.70s %>% 
  select(year) %>% 
  unique()

## Find countries with largest populations as of first year of observations during 1970's
n.ctries <- 3 #number of countries to select
ctries <- gap.70s %>% # Pipe command from the dplyr package can be added with keyboard shortcut command+shift+m
  filter(year == min(yrs)) %>% # Filter to observations from 1972
  arrange(desc(pop)) %>% # Arrange by population
  slice(1:n.ctries) %>% # Choose top n.ctries rows
  select(country) %>% # Select country column
  .$country %>% # Select column
  as.character() # Convert to character vector from factor

```

Using the `filter()` function from the `dplyr` package, we filter the gapminder data to all observations for three countries in the 1970's.  To pick which three countries, we choose countries with the largest populations as of `r min(yrs)`: `r paste(ctries, collapse=", ")`.

```{r displaytibble}
## Create data with 3 countries' observations in the 1970's based on largest populations from first 1970's observation
gap.3 <- gapminder %>% 
   filter(year > 1969,
          year < 1980,
          country %in% ctries) 
kable(gap.3, digits=2)
```

### 1.2 Use Pipe Operator to Select Country and GDP per Capita:

The `select()` function from the `dplyr` package makes it very easy to choose which columns to display from a dataset.  For this question, we select the "country" and "gdpPercap" columns from the filtered dataset presented in 1.1.

```{r pipe}
## Select country and gdpPercap columns from filtered dataset
gap.3 %>% 
  select(country, gdpPercap) %>% 
  kable(digits=2)
```

### 1.3 Filter `gapminder` to observations with drop in Life Expectancy:

We use the `mutate()` function to add a column to the `gapminder` dataset that calculates the change in life expectancy, in years, for each observation.  This column is named "changeinlifeExp".  For each country's first observation, the change is $0$.  Using this new column, we filter out observations whose life expectancy change is negative.  There are  a total of `r gapminder %>% group_by(country) %>% mutate(changeinlifeExp = c(0, diff(lifeExp, lag=1))) %>% filter(changeinlifeExp < 0) %>% nrow()` observations meeting this criteria.  When displaying the life expectancy changes for observations with negative changes, one should note that the values are rounded to two decimal places and a change that displays $0$ is actually a negative change between in the interval $(0, -0.005)$.

```{r LifeExp}
gap.1.3 <- gapminder %>% 
  group_by(country) %>% 
  mutate(changeinlifeExp = c(0, diff(lifeExp, lag=1))) %>% 
  filter(changeinlifeExp < 0) %>% 
  mutate(changeinlifeExp = round(changeinlifeExp, 2)) #round change to 2 digits

kable(gap.1.3[1:10,])
```

### 1.4 Show maximum Country GDP per Capita for each Country:

Using the `group_by()` and `summarise()` functions, this request can be acheived in only a few lines.

```{r maxGDP}
gap.1.4 <- gapminder %>% 
  group_by(country) %>% 
  summarise(maxgdp = round(max(gdpPercap), 2))

kable(gap.1.4[1:10,])
```

### 1.5 Scatterplot of Canada's Life Expectancy vs. GDP per Capita:

By using the `filter()` function, we can pipe the filtered dataset directly into `ggplot()` to create the desired scatterplot.

```{r scatterplot, fig.width=8, fig.height=4}
gapminder %>% 
  filter(country == 'Canada') %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point() + 
  ggtitle('Canada Life Expectancy vs. GDP per Capita') + 
  theme(plot.title = element_text(color='black', face="bold")) +
  scale_x_log10('GDP per Capita on Log Scale', labels=scales::dollar_format()) + 
  ylab('Life Expectancy')
```

## Question 2

We analyze the "continent" categorical variable and the "lifeExp" quantitative varible, which gives the continent for each observation and each observation's life expectancy in years, respectively.  A bar plot for continent shows the possible values of the "continent" variable as well as giving the number of observations from each continent.  The continent with the most observations is `r gapminder %>% group_by(continent) %>% count() %>% arrange(desc(n)) %>% .$continent %>% as.character() %>% .[1]` with `r gapminder %>% group_by(continent) %>% count() %>% .$n %>% max()`.  The continent with the least number of observations is `r gapminder %>% group_by(continent) %>% count() %>% arrange(n) %>% .$continent %>% as.character() %>% .[1]` with `r gapminder %>% group_by(continent) %>% count() %>% .$n %>% min()`.

```{r continent}
ggplot(gapminder, aes(continent)) + 
  geom_bar() + 
  ggtitle('Count of Observations by Continent') + 
  theme(plot.title = element_text(color='black', face="bold")) + 
  ylab('Number of Observations') +
  xlab('Continent')
```

For life expectancy, we show the range of the values and distribution through a histogram, boxplot, and density plot.  We then present summary statistics for life expectancy.  We see the distribution might be bimodal with the mode around 70.  
```{r lifeExp, fig.width=10, fig.height=3}
plot1 <- ggplot(gapminder, aes(lifeExp)) + 
  geom_histogram(bins=50) + 
  ggtitle('Histogram of Life Expectancies') + 
  xlab('Life Expectancy') + 
  ylab('Number of Observations') + 
  theme(plot.title = element_text(color='black', face="bold"))

plot2 <- ggplot(gapminder, aes(y=lifeExp)) + 
  geom_boxplot() + 
  ggtitle('Boxplot of Life Expectancies') + 
  ylab('Life Expectancy') + 
  xlab('') + 
  theme(plot.title = element_text(color='black', face="bold"), 
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot3 <- ggplot(gapminder, aes(lifeExp)) + 
  geom_density() + 
  ggtitle('Density of Life Expectancies') + 
  ylab('Density') + 
  xlab('Life Expectancy') + 
  theme(plot.title = element_text(color='black', face="bold"))

grid.arrange(plot1, plot2, plot3, nrow=1, ncol=3)

summ.le <- summary(gapminder$lifeExp)
names(summ.le) <- c('Minimum', 'First Quartile', 'Median', 'Mean', 'Third Quartile', 'Maximum')
summ.le <- data.frame(rbind(round(summ.le, 2)), row.names = 'Life Expectancy')
kable(summ.le)

```

## Question 3

We now make two plots that can have value for some anlaysis.  The first plot is a scatterplot of GDP per capita by continent versus population-weighted mean life expectancy for each year in the dataset.  This can give us a sense of how life expectancy is changing by continent.  The point of population-weighting the mean life expectancy is so countries with larger populations have more impact on the continent-wide life expectancy.  To calculate GDP per capita, we multiply each country in a continent by its population to get country GDP.  Then by summing country GDP for all countries in a continent, we then divide by the sum of population for all countries on a continent.  This process is repeated for each year in the study.  In order to compare continent life expectancies over time, we use larger plotting symbols for more recent observations.

```{r Q3, fig.width=8, fig.height=4}

## Population by Continent/Year
gapminder %>% 
  group_by(continent, year) %>% 
  mutate(wgt.lifeExp=weighted.mean(lifeExp, w=pop)) %>% 
  mutate(tot.pop=as.numeric(sum(pop)), tot.gdp=as.numeric(sum(pop*gdpPercap))) %>%  #as.numeric() needed for precision
  mutate(tot.gdpPer=round(tot.gdp/tot.pop, 2)) %>% #GDP per capita by continent
  ggplot(aes(x=tot.gdpPer, y=wgt.lifeExp)) + 
  geom_point(aes(shape=continent, color=continent, size=year)) + 
  scale_x_continuous('Continent GDP per Capita', labels=scales::dollar_format()) + 
  ylab('Populated Weighted Mean Life Expectancy') + 
  ggtitle('Continent Level GDP vs Life Expectancy') + 
  theme(plot.title = element_text(color='black', face="bold"))
```

Finally, we compare life expectancies across continents again by use of a boxplot.  Since the distribution of life expectancies may be affected by the number of observations in each continent, we also present a bar plot for the number of countries that have data in each continent.  Thus, while Oceania seems to have a very high life expectancy with low variation, the small number of countries collected in the dataset may not be a true representative sample of the entire continent.

```{r plot2, fig.width=9, fig.height=4}
plot1 <- ggplot(gapminder, aes(x=continent, y=lifeExp)) + 
  geom_boxplot() + 
  xlab('Continent') + 
  ylab('Life Expectancy') + 
  ggtitle('Life Expectancy by Continent') + 
  theme(plot.title = element_text(color='black', face="bold"))

plot2 <- gapminder %>% 
  group_by(continent) %>% 
  mutate(Result=n_distinct(country)) %>% 
  select(continent, Result) %>% 
  unique() %>% 
  ggplot(aes(x=continent, y=Result)) +
  geom_col() +
  ggtitle('Number of Countries from Each Continent') + 
  ylab('Number of Unique Countries') +
  xlab('Continent') + 
  theme(plot.title = element_text(color='black', face="bold"))

grid.arrange(plot1, plot2, nrow=1, ncol=2)
```


## Recycling (Optional):

The results of the code `filter(gapminder, country == c("Rwanda", "Afghanistan"))` are presented below:

```{r recycling, echo=FALSE}
kable(filter(gapminder, country == c("Rwanda", "Afghanistan")))
```

The result of the code is to return 6 entries for Rwanda and Afghanistan.  The analyst is unable to get all of the data for Rwanda and Afghanistan.  The code fetches Rwanda for 1952, Afghanistan for 1957, Rwanda for 1962, etc.  The country fetched for each year in the dataset alternates between Rwanda and Afghanistan.  The correct code to fetch this data is `filter(gapminder, country=="Rwanda" | country=="Afghanistan")`.
```{r correct, echo=FALSE}
kable(filter(gapminder, country=="Rwanda" | country=="Afghanistan")[1:10,])
```