---
title: 'Homework 02'
author: 'Dan Hadley'
output: 
  html_document:
    keep_md: true
    toc: true
editor_options: 
  chunk_output_type: inline
---
<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE,
                      fig.path = "gapminder_files/")
```

**Optional, but recommended startup**:

1. Change the file output to both html and md _documents_ (not notebook).

2. `knit` the document. 

3. Stage and commit the rmd, and knitted documents.

# Intro to `dplyr` syntax

Load the `gapminder` and `tidyverse` packages. Hint: `suppressPackageStartupMessages()`!
    - This loads `dplyr`, too.

<!---The following chunk loads the appropriate libraries--->

```{r load packages, echo=FALSE, warning = FALSE, message = FALSE}
# load your packages here:
suppressPackageStartupMessages(library(gapminder)) #to use the tibble dataset gapminder
suppressPackageStartupMessages(library(tidyverse)) #to load in dplyr package
suppressPackageStartupMessages(library(DT)) #to nicely display tibbles with datatable()
suppressPackageStartupMessages(library(knitr)) #display tables in HTML for R output with kable()
suppressPackageStartupMessages(library(kableExtra)) #options for kable()
```
    
# Question 1

## 1.1 Subset to Three Countries in the 1970's

<!---The following chunk finds the first year of observations for the 70's and the top 3 countries by population for that year--->

```{r subset, echo=FALSE}
## Create data for observations in the 1970's
gap.70s <- gapminder %>% 
  filter(year > 1969,
         year < 1980)

## Create list of years in the 1970's for which there are observations
yrs <- gap.70s %>% 
  select(year) %>% 
  unique()

## Find countries with largest populations as of first year of observations during 1970's
n.ctries <- 3 #number of countries to select
ctries <- gap.70s %>% # Pipe command from the dplyr package can be added with keyboard shortcut command+shift+m
  filter(year == min(yrs)) %>% # Filter to observations from 1972
  arrange(desc(pop)) %>% # Arrange by population
  slice(1:n.ctries) %>% # Choose top n.ctries rows
  select(country) %>% # Select country column
  .$country %>% # Select column
  as.character() # Convert to character vector from factor

```

Using the `filter()` function from the `dplyr` package, we filter the gapminder data to all observations for three countries in the 1970's.  To pick which three countries, we choose countries with the largest populations as of `r min(yrs)`: `r paste(ctries, collapse=", ")`.

```{r displaytibble}
## Create data with 3 countries' observations in the 1970's based on largest populations from first 1970's observation
(gap.3 <- gapminder %>% 
   filter(year > 1969,
          year < 1980,
          country %in% ctries))
```

## 1.2 Use Pipe Operator to Select Country and GDP per Capita:

The `select()` function from the `dplyr` package makes it very easy to choose which columns to display from a dataset.

```{r pipe}
## Select country and gdpPercap columns from filtered dataset
gap.3 %>% 
  select(country, gdpPercap)
```

## 1.3 Filter `gapminder` to observations with drop in Life Expectancy:

First, we can use the `mutate()` function to add a column to the `gapminder` dataset that calculates the 5-year change in life expectancy, in years, for each country.  This column is named `changeinlifeExp`.  For each country's first year observation, the change is 0.  Using this new column, we can filter out observations whose 5-year life expectancy change is negative.  When displaying the negative life expectancy changes, the change values are rounded to two decimal places.  Thus, a change that displays `0` is actually a negative change smaller than `-0.005`.

```{r LifeExp}
gapminder %>% 
  group_by(country) %>% 
  mutate(changeinlifeExp = c(0, diff(lifeExp, lag=1))) %>% 
  filter(changeinlifeExp < 0) %>% 
  mutate(changeinlifeExp = round(changeinlifeExp, 2)) %>% #round change to 2 digits
  datatable()
```

## 1.4 Show maximum Country GDP per Capita for each Country

```{r maxGDP}
gapminder %>% 
  group_by(country) %>% 
  summarise(maxgdp = round(max(gdpPercap), 2)) %>%
  datatable()
```

## 1.5 Scatterplot of Canada's Life Expectancy vs. GDP per Capita

```{r scatterplot, fig.width=8, fig.height=4}
gapminder %>% 
  filter(country == 'Canada') %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point() + 
  ggtitle('Canada') + 
  scale_x_log10('GDP per Capita', labels=scales::dollar_format()) + 
  ylab('Life Expectancy')
```

# Question 2

We analyze the `continent` categorical variable, listing the continent for each observation, and the `lifeExp` quantitative varible, which gives each observation's life expectancy in years.

A bar plot for continent shows the possible values of the `continent` variable as well as giving a sense of the spread of the observations by showing which continents are most/least observed, etc.

```{r continent, fig.cap="Number of observations from each continent"}
ggplot(gapminder, aes(continent)) + 
  geom_bar() + 
  ylab('Number of Observations') +
  xlab('Continent')
```

For life expectancy, 
```{r lifeExp}
ggplot(gapminder, aes(lifeExp)) + 
  geom_histogram(bins=50) + 
  ggtitle('Histogram of Life Expectancies')

ggplot(gapminder, aes(y=lifeExp)) + 
  geom_boxplot() + 
  ggtitle('Boxplot of Life Expectancies')

ggplot(gapminder, aes(lifeExp)) + 
  geom_density()

gapminder %>% 
  select(lifeExp) %>% 
  summary() %>% 
  kable()
```

# Question 3
