---
title: "Assignment 05"
author: "Dan Hadley"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    css: custom.css
  keep_md: true
  toc: true
  toc_float: true
---   

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "hw05_files/")

## Keep track of list of tables
tab.cap = c('Table 1: The factor level orders for continent in the gapminder dataset and continent in the filtered gapminder dataset are the same',
            'Table 2: The factor level orders for continent in the gapminder dataset, the filtered gapminder dataset, and the drop-level gapminder dataset: filtering does not remove the level while droplevels() does',
            'Table 3: Summary of number of observations and levels for the original gapminder dataset, the removal of "Oceania" observations, and dropping the "Oceania" factor level',
            'Table 4: Default ordering of continent factor levels is alphabetical and shown in the left. Factor level ordering changed to descending median regression slope and shown i the right.',
            'Table 5: Data written to csv',
            'Table 6: Data imported from csv')

fig.cap = c('Figure 1: Boxplots of the country slope coefficients for GDP per capita versus year.  The original factor order is alphabetical and the new factor order sorts descendingly by the median value',
            'Figure 2: Plotting country population-weighted average life expectancies for the first 10 country factor levels ordered alphabetically and ordered descendingly by highest population-weighted average life expectancy')
```
  
```{r packages, warning=FALSE, echo=FALSE, message=F}
suppressPackageStartupMessages(library(tidyverse)) #loads ggplot2, tibble, tidyr, dplyr packages
suppressPackageStartupMessages(library(gapminder)) #for gapminder dataset
suppressPackageStartupMessages(library(datasets))
suppressPackageStartupMessages(library(DT)) #for datatable() to display datatables
suppressPackageStartupMessages(library(knitr)) #for kable()
suppressPackageStartupMessages(library(gridExtra)) #arrange figures side-by-side
```

## Introduction

The following section headers come directly from [Assignment 5](https://stat545.stat.ubc.ca/evaluation/hw05/hw05/).  Instructions for each section are given in italics directly below the header as blockquotes.  Note that I have overriden the default size of blockquotes via my `custom.css` file.  Instructions for doing this are taken from [here](https://stackoverflow.com/questions/44191576/how-to-control-the-font-size-of-quoted-text-in-markdown).

---

<br>

---
  
## Exercise 1: Explain the value of the here::here package
  
> *Read through the [blog post](https://malco.io/2018/11/05/why-should-i-use-the-here-package-when-i-m-already-using-projects/) by Malcolm Barrett where he outlines why one should use the here::here package in RStudio projects.*
>
> *Task: * ***In your own words****, summarize the value of the here::here package in 250 words or fewer.*

To explain the usefulness of the `here` package, we first must introduce the idea of `R` projects.  Benefits of using `R` projects are:

  - Divide work into specific projects
  - Uses the project's directory as the working directory
  - Can be set to open a fresh `R` session with an empty global environment and no packages loaded

The `here` package takes advantage of the second bullet point by offering a simple set of functions that detects the project's root directory and works with paths within that directory in a platform-independent way so that the `R` code is not affected by the user's operating system.  Additionally, `here`

  - works when `*.R` files are open in a session outside of a project,
  - gives precedence to a project's directory over a working directory,
  - still allows code to run even when files are moved.

---

<br>

---
  
## Exercise 2: Factor management
  
> *Task: Choose one dataset (of your choice) and a variable to explore.*
  
I work with the `gapminder` dataset and the continent variable, which is set as a factor.

```{r cont-factor}
class(gapminder$continent)
```
  
> *1. Drop Oceania. Filter the Gapminder data to remove observations associated with the `continent` of Oceania. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and Oceania; address the number of rows and the levels of the affected factors.*

Table 1 below contains the summarized results.
  
Removing observations associated with values in a column is easily done via the `dplyr::filter()` command.  We create a new tibble called `noOc` which filters all observations with `continent=="Oceania"` from gapminder.
```{r}
noOc <- gapminder %>% 
  filter(continent != 'Oceania')
```

The number of rows in the `noOc` dataset is `r nrow(noOc)`, as compared to `gapminder` which has `r nrow(gapminder)` rows.  The `continent` column from `gapminder` has `r nlevels(gapminder$continent)` while `noOc` has `r nlevels(noOc$continent)` levels.  The levels of the `continent` factor for both `gapminder` and `noOc` are the same:

```{r levels1}
kable(cbind(levels(gapminder$continent),
            levels(noOc$continent)),
      col.names=c("Gapminder Dataset", "Filtered Dataset"),
      caption=tab.cap[1])
```

We know that "Oceania" is a factor level not included in the data of `noOc`.  To remove it, we use the `droplevels()` command to create a new data frame called `noOc.drop` with the "Oceania" level removed.
```{r drop}
noOc.drop <- noOc %>% 
  droplevels()

kable(cbind(levels(gapminder$continent),
            levels(noOc$continent),
            c(levels(noOc.drop$continent), "")),
      col.names=c("Gapminder Dataset", "Filtered Dataset", "Drop-Level Dataset"),
      caption=tab.cap[2])

```

Note that dropping levels does not change the number of observations.  The number of rows in `noOc` is `r nrow(noOc)` and the number of rows in `noOc.drop` is `r nrow(noOc.drop)`.

In summary, 

Dataset | **`gapminder`** | **`noOc`** | **`noOc.drop`** |
--------|-------------|--------|-------------|
Number of Rows | `r nrow(gapminder)` | `r nrow(noOc)` | `r nrow(noOc.drop)` |
Number of Levels | `r nlevels(gapminder$continent)` | `r nlevels(noOc$continent)` | `r nlevels(noOc.drop$continent)` |
Levels | `r paste(levels(gapminder$continent), collapse=', ')` | `r paste(levels(noOc$continent), collapse=', ')` | `r paste(levels(noOc.drop$continent), collapse=', ')` |
Table: `r tab.cap[3]`


> *2. Reorder the levels of country or continent. Use the forcats package to change the order of the factor levels, based on summarized information of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the mean/median. Use the forcats package in the tidyverse for this, rather than the baseR function as.factor.* 
  
For each country, we calculate the simple linear regression slope of the natural logarithm of gdpPercap against year.  We then reorder the continent factor descendingly by median country-level regression slope coefficient via the `forcats::fct_reorder()` function.  Note that the `forcats` package is automatically loaded into `R` by loading `tidyverse`.
```{r max-gdp, fig.cap=fig.cap[1], fig.width=12, fig.height=4}
# Add column for country regression slope
gap.growth <- gapminder %>% 
  group_by(country) %>% 
  mutate(growth = cor(year, log(gdpPercap))*sd(log(gdpPercap))/sd(year))
  
# Default ordering
p1 <- gap.growth %>% 
  ggplot(aes(x=continent, y=growth)) + 
  geom_boxplot() + 
  xlab("Continent") +
  ylab("Slope") +
  ggtitle(label="Original Factor Order (alphabetical)")

# Order by largest slope
p2 <- gap.growth %>% 
  ggplot(aes(x=fct_reorder(continent, -growth, median), y=growth)) +
  geom_boxplot() + 
  xlab("Continent") +
  ylab("Slope") +
  ggtitle(label="New Factor Order (median)")

# Compare continent factor orderings for boxplots
grid.arrange(p1, p2, nrow=1)
```

```{r compare-levels}
# Change factor level ordering for gap.growth
kable(cbind(levels(gap.growth$continent),
            levels(with(gap.growth, fct_reorder(continent, -growth, median)))),
      col.names=c("Alphabetical (Default)", "Descending by Median Slope"),
      caption=tab.cap[4])
```

---

<br>

---
  
## Exercise 3: File input/output (I/O) 
> **Task: Experiment with at least one of:**
>  
>  - **write_csv()/read_csv() (and/or TSV friends)**,
>  - **saveRDS()/readRDS()**,
>  - **dput()/dget()**.
>  
> **You are expected to create something new, probably by filtering or grouped-summarization of your dataset (for e.g., Singer, Gapminder, or another dataset), export it to disk and then reload it back in using one of the packages above. You should use `here::here()` for reading in and writing out.**
  
Using the gapminder dataset, we calculate the population-weighted moving average life expectancy for each country from `r min(gapminder$year)` to the observation's year.  We then export the country, continent, year, and moving average life expectancy from gapminder to `gap-hw5.csv` using the `readr::write_csv()` function.  The data is imported back into our `R` session via the `readr::read_csv()` function.  The file path is designated by the `here::here()` function.  It should be noted that the `readr` package is automatically loaded with the `tidyverse` package.
```{r read-write}
# Population-weighted moving average function
w.avg <- function(x, w, t) {
  #x is the data to be averaged
  #w is the data by which to weight
  #t is the time column by which we roll the weighted mean
  ans <- numeric(length(x))
  for (i in seq_along(t)) {
    ans[i] <- weighted.mean(x=x[1:i], w=w[1:i])
  }
  return(ans)
}
#Create new column
gap.wavg <- gapminder %>% 
  group_by(country) %>% 
  mutate(wavg.lifeExp = w.avg(x=lifeExp, w=pop, t=year)) %>% 
  select(country, continent, year, lifeExp, wavg.lifeExp)
  
#working directory is project's directory
file.name <- here::here("hw05", "hw05_files", "gap-hw5.csv")

#export csv to disc
write_csv(gap.wavg, file.name)

#re-import csv file
gap.hw5 <- read_csv(file.name) %>% 
  mutate(country = factor(country)) %>% 
  mutate(continent = factor(continent))
```

  
> **With the imported data, play around with factor levels and use factors to order your data with one of your factors (i.e. non-alphabetically).**
  
We reimporting the csv file to the `gap.hw5` tibble, we declared `country` a factor and ordered it alphabetically (the default ordering).  We then reorder the factor levels of `gap.hw5$country` using the `forcats::fct_reorder()` function so that they are ordered descendingly by population-weighted life expectancy for each country as of `r max(gap.hw5$year)`.  Figure 2 shows the factor levels before and after reordering.

```{r country-levels, fig.width=8, fig.height=4, fig.cap=fig.cap[2]}
k1 <- kable(levels(gap.hw5$country)[1:10], col.names="Factor Levels", caption=tab.cap[2])
p1 <- gap.hw5 %>% 
  filter(country %in% levels(country)[1:10]) %>% 
  group_by(country) %>% 
  summarize(last.life = last(wavg.lifeExp)) %>% 
  ggplot(aes(x=country, y=last.life)) +
  geom_point() +
  theme(
    panel.grid.major.x = element_line(colour="grey60", linetype="dashed"),
    axis.text.x = element_text(angle=60)
  ) + 
  ylab("2007 Population-Weighted Life Expectancy") +
  ylim(30, 80) + 
  xlab("Country") +
  ggtitle("First 10 Countries Alphabetically")

p2 <- gap.hw5 %>% 
  mutate(country = fct_reorder(country, -wavg.lifeExp, .fun=last)) %>% 
  filter(country %in% levels(country)[1:10]) %>% 
  group_by(country) %>% 
  summarize(last.life = last(wavg.lifeExp)) %>% 
  ggplot(aes(x=country, y=last.life)) +
  geom_point() +
  theme(
    panel.grid.major.x = element_line(colour="grey60", linetype="dashed"),
    axis.text.x = element_text(angle=60)
  ) + 
  ylab("2007 Population-Weighted Life Expectancy") +
  ylim(30, 80) + 
  xlab("Country") +
  ggtitle("First 10 Countries by Life Expectancy")

grid.arrange(p1, p2, nrow=1)
```

> **For the I/O method(s) you chose, comment on whether or not your newly created file survived the round trip of writing to file then reading back in.**

The newly created file survived the round trip of writin to file and being read back into `R` which can be seen by comparing the two datatables in Table 5 and Table 6.

```{r roundtrip, echo=FALSE}
#Data written to csv
gap.wavg %>% 
  mutate_at(.vars=c("lifeExp", "wavg.lifeExp"), .funs=~round(., digits=2)) %>% 
  datatable(caption=tab.cap[5])

#Data imported from csv
gap.hw5 %>% 
  mutate_at(.vars=c("lifeExp", "wavg.lifeExp"), .funs=~round(., digits=2)) %>% 
  datatable(caption=tab.cap[6])
```

---

<br>

---
  
## Exercise 4: Visualization design
  **Go back through your previous assignments and class participation activities and find figures you created prior to the last week of the course. Recreate at least one figure in light of something you learned in the recent class meetings about visualization design and color.**
  
  **Task: Create a side-by-side plot and juxtapose your first attempt (show the original figure as-is) with a revised attempt after some time spent working on it and implementing principles of effective plotting principles. Comment and reflect on the differences.**
  
  <br/><br/>
  
## Exercise 5: Writing figures to file
  
  **Task: Use ggsave() to explicitly save a plot to file. Include the exported plot as part of your repository and assignment.**