---
title: "Gapminder Exploration"
author: "Dan Hadley"
date: "9/17/2019"
output: 
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
# Set path to save images
knitr::opts_chunk$set(
  fig.path = "hw01_gapminder_files/")
```
```{r load, warning=FALSE, echo=FALSE}
# Load libraries
library(gapminder)
library(tibble)
library(DT)
library(knitr)
```

## 1. Overview

- The `gapminder` dataset contains life expectancy, population, and GDP per capita data for `r length(unique(gapminder$country))` countries

- Data was collected every 5 years from `r min(gapminder$year)` to `r max(gapminder$year)`

- The complete `gapminder` data set has `r nrow(gapminder)` observations.

## 2. Summary

```{r summ, echo=FALSE, results=FALSE}
(summ.gap <- summary(gapminder)) #Surrounding with () declares variable and prints value(s)

cnt <- aggregate(gapminder, by=list(gapminder$continent), length) #counts observations by continent
```

- Africa has `r max(cnt$continent)` observations, `r diff(sort(cnt$continent)[1:2])` more than the next continent, `r cnt$Group.1[which(rank(-cnt$continent) == 2)]`

- Across all countries, GDP per capita ranges from `r min(cnt$gdpPercap)` USD to `r max(cnt$gdpPercap)` USD

- The mean (`r mean(gapminder$gdpPercap)`) is twice as large as the median (`r median(gapminder$gdpPercap)`)

- GDP per capita data has a large positive skew.

## 3. Life Expectancy by Continent

Life expectancies of Asian countries have steadily increased

```{r years, echo=FALSE}
yrs <- unique(gapminder$year)
yrs.seq <- yrs[c(1, 4, 8, 12)]

par(mfrow=c(2,2))
for (i in yrs.seq) {
  dat <- gapminder[gapminder$year == i,]
  boxplot(lifeExp ~ continent, data=dat,
          main=paste("Life Expectancies in", i),
          xlab="Continent", ylab="Years",
          ylim=c(30, 82))
}
par(mfrow=c(1,1))
```


## 4. Life Expectancy and GDP per Capita in 2007

- Longest life expectancy: Japan (`r round(max(dat$lifeExp), 1)` years)
- Highest GDP per capita: Norway (US\$`r format(max(dat$gdpPercap), scientific=FALSE)`)
- Shortest life expectancy: Swaziland (`r round(min(dat$lifeExp), 1)` years)

```{r gdp-vs-lifeExp, echo=FALSE, fig.width=10, fig.height=6}
dat <- gapminder[gapminder$year == max(gapminder$year), c(1, 4, 6)]

plot(dat$gdpPercap, dat$lifeExp, pch=16, cex=0.5,
     xlab="GDP per Capita", ylab="Life Expectancy",
     main='Plot of 2007 Life Expectancy against 2007 GDP per Capita',
     ylim=c(30, 90))
points(x=c(dat$gdpPercap[which.max(dat$lifeExp)], 
           max(dat$gdpPercap),
           dat$gdpPercap[which.min(dat$lifeExp)]),
       y=c(max(dat$lifeExp),
           dat$lifeExp[which.max(dat$gdpPercap)],
           min(dat$lifeExp)),
       pch=16, cex=0.75, col='red')
text(x=dat$gdpPercap[which.max(dat$lifeExp)], y=max(dat$lifeExp)+2,
     labels=dat$country[which.max(dat$lifeExp)], cex=0.95)
text(x=max(dat$gdpPercap), y=dat$lifeExp[which.max(dat$gdpPercap)]+2,
     labels=dat$country[which.max(dat$gdpPercap)], cex=0.95)
text(x=dat$gdpPercap[which.min(dat$lifeExp)], y=min(dat$lifeExp)-2,
     labels=dat$country[which.min(dat$lifeExp)], cex=0.95)

```

## 5. Life Expectancy and GDP per Capita for Canada

- Life expectancy and GDP per capita for Canada have both increased over time
- The relationship has been approximately linear

```{r Canada, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
can <- gapminder[gapminder$country=='Canada',] #Subset gapminder dataset to Canada's observations

# Plot with year as plot symbol
plot(can$gdpPercap, can$lifeExp, type='n',
     xlab="GDP per Capita", ylab="Life Expectancy",
     main="Data for Canada from 1952 thru 2007")
text(can$gdpPercap, can$lifeExp, labels=can$year, cex=0.9)
```