---
title: "Exploration of Gapminder"
author: "Dan Hadley"
date: "9/17/2019"
output: 
  github_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
# Set path to save images
knitr::opts_chunk$set(
  fig.path = "hw01_gapminder_files/")
```

```{r load, warning=FALSE, echo=FALSE}
# Load libraries
library(gapminder)
library(tibble)
library(DT)
library(knitr)
```

## 1. Introduction

We explore the gapminder dataset available in the gapminder package.  All analyses are performed using the `R` software with the following packages: [gapminder](https://cran.r-project.org/web/packages/gapminder/index.html), [tibble](https://cran.r-project.org/web/packages/tibble/index.html), [DT](https://cran.r-project.org/web/packages/DT/index.html), [knitr](https://cran.r-project.org/web/packages/knitr/index.html).
<!--[ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html),-->

This exploration should give the reader an idea of the type of data the `gapminder` dataset contains and some interesting analyses that can be performed on the data.

## 2. Overview

To start, we can view the column names in the dataset using the `names()` function.
```{r colnames, echo=FALSE}
names(gapminder)
```

The columns in the dataset are Country, Continent, Year, Life Expectancy, Population, and GDP per capita.  To see the data type for each column, we can view the structure of the dataset via the `str()` function.
```{r structure, echo=FALSE}
str(gapminder)
```

The result shows the `gapminder` class as well as the data type for each column.  Note that both country and continent are labeled as factors.  Factor is a data type in `R` used for character data, essentially treating each unique character entry as a group.  The data type of each column is useful when applying functions to column data.


To get a feel for the gapminder dataset, we can view the first 6 rows using the `head()` function.
```{r, echo=FALSE}
head(gapminder)
```

The result is a 6 row and 6 column object called a tibble.  The dataset contains information on the life expectancy (in years), population, and GDP per capita (US\$, inflation-adjusted) for `r length(unique(gapminder$country))` countries from `r min(gapminder$year)` to `r max(gapminder$year)` measured every 5 years.  The complete `gapminder` data set has `r nrow(gapminder)` observations.

## 3. Analysis

To get a feel for the dataset, the `summary()` function calclulates the minimum, 3rd quartile, median, mean, 1st quartile, and maximum value for all numeric columns.  Columns with character data, often saved as the factor data type in `R`, performs counts on the unique character entries.
```{r summ, echo=FALSE}
(summ.gap <- summary(gapminder)) #Surrounding with () declares variable and prints value(s)

cnt <- aggregate(gapminder, by=list(gapminder$continent), length) #counts observations by continent
```

Africa has `r max(cnt$continent)` observations, which is `r diff(sort(cnt$continent)[1:2])` more than the next continent, `r cnt$Group.1[which(rank(-cnt$continent) == 2)]`.  For numerical data, `summary()` presents a quick snapshot of its distribution.  For example, we see that GDP per capita ranges from `r min(cnt$gdpPercap)` USD to `r max(cnt$gdpPercap)` USD, and the mean (`r mean(gapminder$gdpPercap)`) is twice as large as the median (`r median(gapminder$gdpPercap)`).  Thus, the GDP per capita data has a large positive skew.

### 3.1 Life Expectancy by Continent

```{r years, echo=FALSE}
yrs <- unique(gapminder$year)
yrs.seq <- yrs[c(1, 4, 8, 12)]
```

Like `summary()`, boxplots provide a snapshot of the distribution of data, allowing comparison across groups.  In this case, we compare the distribution of life expectancies by continent for four years: `r combine_words(yrs.seq)`.

```{r boxplots, echo=FALSE, fig.width=10, fig.height=7}
par(mfrow=c(2,2))
for (i in yrs.seq) {
  dat <- gapminder[gapminder$year == i,]
  boxplot(lifeExp ~ continent, data=dat,
          main=paste("Life Expectancies in", i),
          xlab="Continent", ylab="Years",
          ylim=c(30, 82))
}
par(mfrow=c(1,1))
```

From the boxplots, it is clear that life expectancies of Asian countries have steadily increased since 1952, and the interquartile range has gotten much smaller.  This indicates that the difference between countries with long and short life expectancies is shrinking.

### 3.2 Life Expectancy and GDP per Capita in 2007

The relationship between Life Expectancy and GDP per capita for the most recent year, `r max(gapminder$year)`, can be visualized with a scatterplot for all `r nrow(gapminder[gapminder$year == max(gapminder$year),])` countries.

```{r gdp-vs-lifeExp, echo=FALSE}
dat <- gapminder[gapminder$year == max(gapminder$year), c(1, 4, 6)]

plot(dat$gdpPercap, dat$lifeExp, pch=16, cex=0.5,
     xlab="GDP per Capita", ylab="Life Expectancy",
     main='Plot of 2007 Life Expectancy against 2007 GDP per Capita',
     ylim=c(30, 90))
text(x=dat$gdpPercap[which.max(dat$lifeExp)], y=max(dat$lifeExp)+2,
     labels=dat$country[which.max(dat$lifeExp)], cex=0.75)
text(x=max(dat$gdpPercap), y=dat$lifeExp[which.max(dat$gdpPercap)]+2,
     labels=dat$country[which.max(dat$gdpPercap)], cex=0.75)
text(x=dat$gdpPercap[which.min(dat$lifeExp)], y=min(dat$lifeExp)-2,
     labels=dat$country[which.min(dat$lifeExp)], cex=0.75)

```

Japan had the highest life expectancy in 2007 of `r round(max(dat$lifeExp), 1)` years, and Norway had the highest GDP per capita in 2007 at US\$`r format(max(dat$gdpPercap), scientific=FALSE)`.  Meanwhile, Swaziland had the lowest life expectancy in 2007 at `r round(min(dat$lifeExp), 1)` years.

### 3.3 Life Expectancy and GDP per Capita for Canada

The relationship between life expectancy and GDP per capita for Canada over time can be visualized by a scatterplot of the data for every 5 years from `r min(gapminder$year)` to `r max(gapminder$year)`. The plot symbols are the years of the observations, which indicate an increasing relationship over time.

```{r Canada, echo=FALSE, warning=FALSE, fig.cap="Plot of Life Expectancy (years) versus GDP per capita (USD) in Canada every five years from 1952 to 2007"}
can <- gapminder[gapminder$country=='Canada',] #Subset gapminder dataset to Canada's observations

# Plot with year as plot symbol
plot(can$gdpPercap, can$lifeExp, type='n',
     xlab="GDP per Capita", ylab="Life Expectancy",
     main="Data for Canada from 1952 thru 2007")
text(can$gdpPercap, can$lifeExp, labels=can$year, cex=0.8)
```